An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-21 paper.tex
Run spell check

We can analyze the spread of information between $A$, $B$, and $C$
as a process on a graph that changes over time.
Such graphs are known as evolving graphs.
% Evolving graphs are models to study this kind of scenario where the
% interrelations between entities are time-dependent.
% We assume that the reader is familiar with the basic graph terminology.
% All the necessary material can be found in}.
We follow the notation of (static) graphs in \cite{even12,kegi11}.
In general, a evolving graph $G(t) = (V(t), E(t))$ is  a graph where the node set
and edge set are functions of time $t$.
However, in different literatures the term ``evolving graph'' has been
used to refer to slightly different models.
 Flajolet et al. \cite{fkp89} use the term to mean the evolution of random graphs.
Bahmani et al. \cite{bkmu12} use the term to mean a sequence of graphs where
the difference between two consecutive graphs is small.
Similar models are also investigated using the term ``evolving network'' and
``time-dependent graph''.

In this work, we consider a discretization of the timeline as $n$ distinct
timestamps and define an evolving graph $G_n$ to be
a sequence of (static) graphs
$G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$.
Depends on applications, the graph $G^{[i]}$ could mean either
a snapshot of $G(t)$ at timestamp $i$ or  the union of all the graphs
between timestamp $i-1$ and $i$ (excludes timestamp $i-1$).
Both the node set and edge set of $G^{[i]}$ could be different from
these of $G^{[j]}, j \ne i$.
\begin{center}
\begin{tikzpicture}[snake=zigzag, line before snake = 5mm, line after snake = 5mm]
    % draw horizontal line
    \draw (0,0) -- (2,0);
    \draw[snake] (2,0) -- (4,0);
    \draw (4,0) -- (5,0);
    \draw[snake] (5,0) -- (7,0);

@@

    % draw nodes
    \draw (0,0) node[below=3pt] {$ 1 $} node[above=3pt] {$   $};
    \draw (1,0) node[below=3pt] {$ 2 $} node[above=3pt] {$ G^{[2]} $};
    \draw (2,0) node[below=3pt] {$ 3 $} node[above=3pt] {$  $};
    \draw (3,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (4,0) node[below=3pt] {$ 5 $} node[above=3pt] {$  $};
    \draw (5,0) node[below=3pt] {$ 6 $} node[above=3pt] {$ G^{[6]}$};
    \draw (6,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (7,0) node[below=3pt] {$ n $} node[above=3pt] {$ $};
  \end{tikzpicture}
\end{center}
Our model is the same as that of Tang et al. \cite{tmml10} though they
use slightly different notation.
Grindrod et al.'s model  \cite{gphe11} \cite{grihig13} is very similar to ours,
where an evolving graph is considered as a sequence of adjacency matrices.
Our model can also be seen as a special case of
the multilayer networks discussed by Kivel{\"a} et al. \cite{kabg14}, where
there is only one set of layers for time.
% We present a preliminary discussion of using the model to mine citation networks.

@@

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{An evolving graph with 3 timestamps $t_1$, $t_2$ and $t_3$.
At each timestamp, the evolving graph is represented as a graph.
The green filled nodes in the figure are active nodes.}
\label{fig:eg_shortest_path}
\end{figure}

We show a small evolving graph $ \langle G^{[1]}, G^{[2]}, G^{[3]}\rangle$ with $3$ timestamps $t_1$, $t_2$ and $t_3$
in Figure \ref{fig:eg_shortest_path}, where each $G^{[i]}$ is a graph at timestamp $t_i$.
For example,
at timestamp $t_1$, the graph $G^{[1]}$ contains three nodes $1$, $2$ and $3$ and $(1,2)$ is an edge of $G^{[1]}$.
Though Figure \ref{fig:eg_shortest_path} presents a
directed evolving graph,  the algorithms that we shall discuss work on both
directed and undirected evolving graphs.
We shall use this example throughout the paper to illustrate various
notions.

Formally, an evolving graph $G_n $ is defined to be a sequence of graphs $\langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$,  where each $G^{[t]} = (V^{[t]}, E^{[t]})$ represents a graph at timestamp $t$.
We shall use the notation $(v, t)$ to represent
a node $v$ at timestamp $t$.
We say two nodes are \emph{connected} if there is an edge between them
and we say a node $(v,t)$ is \emph{active}, if $v$ is connected to at least a node other than itself at timestamp $t$. Otherwise we say it is \emph{inactive}.
In Figure \ref{fig:eg_shortest_path}, node $(1,t_1)$ and $(2,t_2)$  are active but node $(3,t_1)$ is inactive.

In order to define the evolving-graph BFS, we need the concept of a
\emph{temporal path}.
A temporal path on an evolving graph $G_n$ is a sequence of node-timestamp
pair $\langle (v_1, t_1), (v_2, t_2), \ldots, (v_m, t_m) \rangle$ such that
$t_1 \leq t_2 \leq \cdots \leq t_m$ and $v_i = v_j$ if and only if
$t_i \ne t_j$. The length of a temporal path is the length of the sequence.
For example, $\langle (1, t_1), (1, t_2), (3, t_2), (3, t_3) \rangle$ and
$\langle (1, t_1), (2, t_1), (2, t_3), (3, t_3) \rangle$
are two temporal paths from $(1, t_1)$ to $(3, t_3)$
on the evolving graph in Figure \ref{fig:eg_shortest_path}.

We would like a BFS on evolving graphs to work similar to that on graphs: the search starts with some source node $(s,t)$, called \emph{root}. It constructs a tree through discovering all nodes that are reachable by
temporal paths of length $k$ before moving to all nodes that are reachable by
 temporal paths of length $k+1$.
By construction, the timestamp of an active node is always not later than
that of its descendant. Suppose we
start with node $(1, t_1)$ in Figure \ref{fig:eg_shortest_path}.
After $1$ temporal path from $(1, t_1)$, we can reach
$(2, t_1)$ and $(1, t_2)$. After $2$ temporal paths, we can reach
$(2, t_1), (1, t_2), (2, t_2)$ and $(3, t_2)$.
% This example motivates us to define the reachable set of
% a node-timestamp pair.
Formally, we define the \emph{forward-neighbors} of a node $(v, t)$ to
be the set of all the reachable nodes after $1$ temporal path from
$(v, t)$, i.e., the nodes pointed by $v$ at timestamp $t$ and
the active node $v$ itself at later timestamps. In Figure
\ref{fig:eg_shortest_path},
the forward-neighbors of $(1, t_1)$  are $(2, t_1)$ and $(1, t_2)$ and
the forward-neighbors of $(2, t_1)$ is $(2, t_3)$.
See Algorithm  \ref{alg:bfs} for the evolving-graph BFS algorithm.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{BFS($G_n, (s,t)$)}{
  $dict.(s,t) = 0$ \\
  $i = 1$ \\
  $fronter = \emptyset$ \\
  \enqueue{$frontier, (s,t)$} \\
 \While{$frontier \ne \emptyset$}{
   $next = \emptyset$ \\
   \For{$(u,t') \in frontier$}{
     \For{$(v,t'') \in$\nei{$(u,t')$}}{
          \If{$(v,t'') \notin d$}{
              $dict.(v,t'') = i$\\
              \enqueue{$next, (v,t'')$}
              }
         }
       }
  $frontier = next$ \\
  $i = i+1$ \\
  }
  \textbf{return} $dict$
}
 \caption{Evolving-graph BFS.
Let $G_n$ be an
evolving graph represented by adjacency lists and $(s, t)$ is a node
of $G_n$.
For each node-timestamp pair $(v_i,t_i)$,
we use a dictionary $dict.(v_i,t_i)$ to store its distance (the length of temporal
paths) from $(s, t)$. We also use
two first-in, first-out queue $frontier$ and $next$, where $frontier$
stores the frontier nodes of the current discovered regions and $next$ stores
the frontier nodes of the next level.
The algorithm returns $dict$ which stores all reachable node-timestamp pairs from $(s,t)$ such that if $dict.(v_i, t_i) = k$,
the length of the shortest temporal path
from $(s,t)$ to $(v_i, t_i)$ is $k$.}
 \label{alg:bfs}
\end{algorithm}

@@

\subsection{Algebraic Evolving-Graph BFS}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ timestamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) =
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$.
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as
\[
\Big\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\Big\rangle
\]

This algebraic formulation exploits a graphical interpretation of matrix--vector
products involving the adjacency matrix~\cite{kegi11}.
Let $G = (V, E)$ be a (directed or undirected) graph
and $k$ be a node in the graph $G$. If $A$ is the adjacency matrix of $G$ and
$b$ is the column vector
\[
b_i =
\begin{cases}
1 & \mbox{if $i = k$,} \\
0 & \mbox{otherwise,}
\end{cases}
\]
then $A^Tb$ is a column vector where the indices of the nonzero entries
indicate the forward-neighbors of $k$.

The evolving graph BFS can be thought of as a generalization of blocked
matrix--vector products.
Introduce a new operator $\odot$, a variation on the ordinary matrix--vector
product such that
\[
A \odot b =
\begin{cases}
b & \mbox{if $Ab \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
Then, the forward-neighbors of node $(k, 1)$ in $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ are the nonzero indices of the
elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Tb, \quad  (A^{[2]})^T\odot b, \quad \ldots \quad (A^{[n]})^T\odot b \big\rangle.
\end{equation}
The forward-neighbors at timestamp $t$ are the nonzero indices of the $t$-th element of the above sequence.
Hence given $A_n$ and a vector $b$, we can use \eqref{eq:out_ne} to
get a $1$-step BFS tree. In Figure \ref{fig:eg_shortest_path}, the forward-neighbors
of node $(1, t_1)$ can be computed by
\[
\Big\langle
  \begin{bmatrix}
    0 & 0 & 0 \\
    1 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix}
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
,
 \begin{bmatrix}
   0 & 0 & 0 \\
   0 & 0 & 0 \\
   1 & 0 & 0
 \end{bmatrix}
\odot
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
,
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 0 \\
  0 & 1 & 0
 \end{bmatrix}
\odot
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
\Big\rangle
\]
which is equal to
\[
\Big\langle
\begin{bmatrix}
0 \\
1 \\
0
\end{bmatrix},~
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix},~
\begin{bmatrix}
0 \\
0 \\
0
\end{bmatrix}
\Big\rangle
\]
This means node $(2,t_1)$ and $(1,t_2)$ are the forward-neighbors of $(1,t_1)$.

Let $(A^{[i]})^Tb = c_i$.
At the second step, we need to compute the following components
\begin{align}
  \label{eq:5}
 \big\langle (A^{[1]})^Tc_1,   (A^{[2]})^T\odot c_1, \ldots,  (A^{[n]})^T\odot c_1  \big\rangle  \\
\big\langle (A^{[2]})^Tc_{2}, \ldots, (A^{[n]})^T\odot c_{2} \big\rangle \\
 \ldots  \\
  \big\langle (A^{[n]})^Tc_n \big\rangle
\label{eq:6}
\end{align}
If we sum these resulting vectors according to timestamps, we obtain
the forward-neighbors of the nodes computed at step $1$.
For $n =2$, we have
\[
\Big\langle
(A^{[1]})^Tc_1,~
(A^{[2]})^T\odot c_1 + (A^{[2]})^T c_2
\Big\rangle.
\]
If we form a block triangular matrix $\bm{A_n}^T$
\[
\bm{A_n}^T =
\begin{bmatrix}
(A^{[1]})^T &   0            & \cdots  &  & 0 \\
\sigma(A^{[2]})^T & (A^{[2]})^T & 0         &\cdots  & \vdots \\
\sigma(A^{[3]})^T & \sigma(A^{[3]})^T & (A^{[3]})^T & \cdots \\
   \vdots              &                              &      \vdots            &  &  \vdots \\
\sigma(A^{[n]})^T &                  &       \cdots                      &  & (A^{[n]})^T
\end{bmatrix},
\]
where  $\sigma(A^{[i]})^Tb = (A^{[i]})^T\odot b$, and a block vector $\bm{b}$
\[
\bm{b}^T = [b^T, 0, \cdots, 0  ],
\]
then \eqref{eq:out_ne} is equivalent to $\bm{A_n}^T\bm{b}$ and
summing \eqref{eq:5}-\eqref{eq:6} is equivalent to $\bm{A_n}^T(\bm{A_n}^T\bm{b})$.
In general, we can define an evolving-graph BFS using
block vector-matrix multiplications. The algorithm is presented in Algorithm
\ref{alg:bfsla}.

@@

% adjacency matrices
Suppose $G_n$ is represented by adjacency lists.
Then the number of adjacency lists is the same as the number
of active nodes in $G_n$. So for large $n$, this representation
can use a large amount of memory.
For the graph $G^{[i]} = (V^{[i]}, E^{[i]})$ at timestamp $i$,
we denote the active node set of $V^{[i]}$ as $\tilde V^{[i]}$,
where we have $\tilde V^{[i]} \subseteq V^{[i]}$.
Step $10$ of  Algorithm \ref{alg:bfs} ensures each active node-timestamp pair in $G_n$ is enqueued in $next$ (and then $frontier$) at most once.
And so \texttt{forward-neighbors} loops through each
active node-timestamp pair at most once.
The operations of enqueuing take $O(1)$ time.
Thus the total operations of enqueuing are
$O(\sum_{i=1}^n|\tilde V^{[i]}|)$, where $|\cdot|$ represents the number of elements of a set.  Let $\sum_i |\tilde V^{[i]}| = |\tilde V|$ and $\sum_i |E^{[i]}| = |\hat E|$.
Then the time of scanning
\texttt{forward-neighbors} is
\[
\sum_{i=1}^n\sum_{v \in \tilde V^{[i]}} \texttt{forward-neighbors}(v) =
O(|\hat E|),
\]
by the definition of forward-neighbors.
Hence the total running time is $O(|\hat E|+ |\tilde V|)$.

@@

Given a root node $(s, t)$, during the execution the BFS discussed above builds a search tree, denoted by $T(s,t)$,
where nodes are temporally connected.
\begin{lemma}
 For a node active at different timestamps, say $(s, t_1)$ and $(s,t_2)$ such that $t_1 < t_2$,  we have $T(s, t_2) \subset T(s, t_1)$.
\end{lemma}
\proof
Suppose $(v, t)$ belongs to $T(s, t_2)$. Then there exits an temporal path
(constructed by the evolving graph BFS) from $(s,t_2)$ to $(v,t)$. Since both
$(s,t_1)$ and $(s,t_2)$ are active, $(s,t_2)$ is an out-neighbor of $(s,t_1)$ by definition, i.e., there exits an edge from $(s,t_1)$ to $(s,t_2)$.  Hence there exits an temporal path
from $(s,t_1)$ to $(v,t)$ and so $(v,t) \in T(s,t_1)$.
\endproof

>>>>

We can analyze the spread of information between $A$, $B$, and $C$
as a process on a graph that changes over time.
Such graphs are known as evolving graphs.
% Evolving graphs are models to study this kind of scenario where the
% interrelations between entities are time-dependent.
% We assume that the reader is familiar with the basic graph terminology.
% All the necessary material can be found in}.
We follow the notation of (static) graphs in \cite{even12,kegi11}.
In general, a evolving graph $G(t) = (V(t), E(t))$ is  a graph where the node set
and edge set are functions of time $t$.
However, in different literature the term ``evolving graph'' has been
used to refer to slightly different models.
 Flajolet et al. \cite{fkp89} use the term to mean the evolution of random graphs.
Bahmani et al. \cite{bkmu12} use the term to mean a sequence of graphs where
the difference between two consecutive graphs is small.
Similar models are also investigated using the term ``evolving network'' and
``time-dependent graph''.

In this work, we consider a discretization of the time line as $n$ distinct
time stamps and define an evolving graph $G_n$ to be
a sequence of (static) graphs
$G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$.
Depends on applications, the graph $G^{[i]}$ could mean either
a snapshot of $G(t)$ at time stamp $i$ or  the union of all the graphs
between time stamp $i-1$ and $i$ (excludes time stamp $i-1$).
Both the node set and edge set of $G^{[i]}$ could be different from
these of $G^{[j]}, j \ne i$.
\begin{center}
\begin{tikzpicture}[snake=zigzag, line before snake = 5mm, line after snake = 5mm]
    % draw horizontal line
    \draw (0,0) -- (2,0);
    \draw[snake] (2,0) -- (4,0);
    \draw (4,0) -- (5,0);
    \draw[snake] (5,0) -- (7,0);

@@

    % draw nodes
    \draw (0,0) node[below=3pt] {$ 1 $} node[above=3pt] {$   $};
    \draw (1,0) node[below=3pt] {$ 2 $} node[above=3pt] {$ G^{[2]} $};
    \draw (2,0) node[below=3pt] {$ 3 $} node[above=3pt] {$  $};
    \draw (3,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (4,0) node[below=3pt] {$ 5 $} node[above=3pt] {$  $};
    \draw (5,0) node[below=3pt] {$ 6 $} node[above=3pt] {$ G^{[6]}$};
    \draw (6,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (7,0) node[below=3pt] {$ n $} node[above=3pt] {$ $};
  \end{tikzpicture}
\end{center}
Our model is the same as that of Tang et al. \cite{tmml10} though they
use slightly different notation.
Grindrod et al.'s model  \cite{gphe11} \cite{grihig13} is very similar to ours,
where an evolving graph is considered as a sequence of adjacency matrices.
Our model can also be seen as a special case of
the multi-layer networks discussed by Kivel{\"a} et al. \cite{kabg14}, where
there is only one set of layers for time.
% We present a preliminary discussion of using the model to mine citation networks.

@@

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{An evolving graph with 3 time stamps $t_1$, $t_2$ and $t_3$.
At each time stamp, the evolving graph is represented as a graph.
The green filled nodes in the figure are active nodes.}
\label{fig:eg_shortest_path}
\end{figure}

We show a small evolving graph $ \langle G^{[1]}, G^{[2]}, G^{[3]}\rangle$ with $3$ time stamps $t_1$, $t_2$ and $t_3$
in Figure \ref{fig:eg_shortest_path}, where each $G^{[i]}$ is a graph at time stamp $t_i$.
For example,
at time stamp $t_1$, the graph $G^{[1]}$ contains three nodes $1$, $2$ and $3$ and $(1,2)$ is an edge of $G^{[1]}$.
Though Figure \ref{fig:eg_shortest_path} presents a
directed evolving graph,  the algorithms that we shall discuss work on both
directed and undirected evolving graphs.
We shall use this example throughout the paper to illustrate various
notions.

Formally, an evolving graph $G_n $ is defined to be a sequence of graphs $\langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$,  where each $G^{[t]} = (V^{[t]}, E^{[t]})$ represents a graph at time stamp $t$.
We shall use the notation $(v, t)$ to represent
a node $v$ at time stamp $t$.
We say two nodes are \emph{connected} if there is an edge between them
and we say a node $(v,t)$ is \emph{active}, if $v$ is connected to at least a node other than itself at time stamp $t$. Otherwise we say it is \emph{inactive}.
In Figure \ref{fig:eg_shortest_path}, node $(1,t_1)$ and $(2,t_2)$  are active but node $(3,t_1)$ is inactive.

In order to define the evolving-graph BFS, we need the concept of a
\emph{temporal path}.
A temporal path on an evolving graph $G_n$ is a sequence of node-time stamp
pair $\langle (v_1, t_1), (v_2, t_2), \ldots, (v_m, t_m) \rangle$ such that
$t_1 \leq t_2 \leq \cdots \leq t_m$ and $v_i = v_j$ if and only if
$t_i \ne t_j$. The length of a temporal path is the length of the sequence.
For example, $\langle (1, t_1), (1, t_2), (3, t_2), (3, t_3) \rangle$ and
$\langle (1, t_1), (2, t_1), (2, t_3), (3, t_3) \rangle$
are two temporal paths from $(1, t_1)$ to $(3, t_3)$
on the evolving graph in Figure \ref{fig:eg_shortest_path}.

We would like a BFS on evolving graphs to work similar to that on graphs: the search starts with some source node $(s,t)$, called \emph{root}. It constructs a tree through discovering all nodes that are reachable by
temporal paths of length $k$ before moving to all nodes that are reachable by
 temporal paths of length $k+1$.
By construction, the time stamp of an active node is always not later than
that of its descendant. Suppose we
start with node $(1, t_1)$ in Figure \ref{fig:eg_shortest_path}.
After $1$ temporal path from $(1, t_1)$, we can reach
$(2, t_1)$ and $(1, t_2)$. After $2$ temporal paths, we can reach
$(2, t_1), (1, t_2), (2, t_2)$ and $(3, t_2)$.
% This example motivates us to define the reachable set of
% a node-timestamp pair.
Formally, we define the \emph{forward-neighbors} of a node $(v, t)$ to
be the set of all the reachable nodes after $1$ temporal path from
$(v, t)$, i.e., the nodes pointed by $v$ at time stamp $t$ and
the active node $v$ itself at later time stamps. In Figure
\ref{fig:eg_shortest_path},
the forward-neighbors of $(1, t_1)$  are $(2, t_1)$ and $(1, t_2)$ and
the forward-neighbors of $(2, t_1)$ is $(2, t_3)$.
See Algorithm  \ref{alg:bfs} for the evolving-graph BFS algorithm.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{BFS($G_n, (s,t)$)}{
  $dict.(s,t) = 0$ \\
  $i = 1$ \\
  $fronter = \emptyset$ \\
  \enqueue{$frontier, (s,t)$} \\
 \While{$frontier \ne \emptyset$}{
   $next = \emptyset$ \\
   \For{$(u,t') \in frontier$}{
     \For{$(v,t'') \in$\nei{$(u,t')$}}{
          \If{$(v,t'') \notin d$}{
              $dict.(v,t'') = i$\\
              \enqueue{$next, (v,t'')$}
              }
         }
       }
  $frontier = next$ \\
  $i = i+1$ \\
  }
  \textbf{return} $dict$
}
 \caption{Evolving-graph BFS.
Let $G_n$ be an
evolving graph represented by adjacency lists and $(s, t)$ is a node
of $G_n$.
For each node-time stamp pair $(v_i,t_i)$,
we use a dictionary $dict.(v_i,t_i)$ to store its distance (the length of temporal
paths) from $(s, t)$. We also use
two first-in, first-out queue $frontier$ and $next$, where $frontier$
stores the frontier nodes of the current discovered regions and $next$ stores
the frontier nodes of the next level.
The algorithm returns $dict$ which stores all reachable node-time stamp pairs from $(s,t)$ such that if $dict.(v_i, t_i) = k$,
the length of the shortest temporal path
from $(s,t)$ to $(v_i, t_i)$ is $k$.}
 \label{alg:bfs}
\end{algorithm}

@@

\subsection{Algebraic Evolving-Graph BFS}
\label{sec:representation}
As before, let $G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]}\rangle$ be an evolving graph with $n$ time stamps. For each graph $G^{[t]}$, we use $A^{[t]}$ to denote
the corresponding adjacency matrix.
\[
A^{[t]}(i,j) =
\begin{cases}
 1 & \mbox{if $(i,j)$ is an edge of $G^{[t]}$,} \\
 0 & \mbox{otherwise.}
\end{cases}
\]
 Each $A^{[t]}$ can be represented by
a $|V|\times |V|$ sparse matrix,
where $|V|$ is the total number of nodes in $G_n$.
We use a sequence of adjacency matrices $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ to represent $G_n$.
For example, the evolving graph in Figure \ref{fig:eg_shortest_path}
can be represented as
\[
\Big\langle
  \begin{bmatrix}
    0 & 1 & 0 \\
    0 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix},~
 \begin{bmatrix}
   0 & 0 & 1 \\
   0 & 0 & 0 \\
   0 & 0 & 0
 \end{bmatrix},~
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 1 \\
  0 & 0 & 0
 \end{bmatrix}
\Big\rangle
\]

This algebraic formulation exploits a graphical interpretation of matrix--vector
products involving the adjacency matrix~\cite{kegi11}.
Let $G = (V, E)$ be a (directed or undirected) graph
and $k$ be a node in the graph $G$. If $A$ is the adjacency matrix of $G$ and
$b$ is the column vector
\[
b_i =
\begin{cases}
1 & \mbox{if $i = k$,} \\
0 & \mbox{otherwise,}
\end{cases}
\]
then $A^Tb$ is a column vector whose nonzero entries have indices that are the
forward-neighbors of $k$.

The evolving graph BFS can be thought of as a generalization of blocked
matrix--vector products.
Introduce a new operator $\odot$, a variation on the ordinary matrix--vector
product such that
\[
A \odot b =
\begin{cases}
b & \mbox{if $Ab \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
Then, the forward-neighbors of node $(k, 1)$ in $A_n = \langle A^{[1]}, A^{[2]}, \ldots, A^{n]}\rangle$ are the nonzero indices of the
elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Tb, \quad  (A^{[2]})^T\odot b, \quad \ldots \quad (A^{[n]})^T\odot b \big\rangle.
\end{equation}
The forward-neighbors at time stamp $t$ are the nonzero indices of the $t$-th element of the above sequence.
Hence given $A_n$ and a vector $b$, we can use \eqref{eq:out_ne} to
get a $1$-step BFS tree. In Figure \ref{fig:eg_shortest_path}, the forward-neighbors
of node $(1, t_1)$ can be computed by
\[
\Big\langle
  \begin{bmatrix}
    0 & 0 & 0 \\
    1 & 0 & 0 \\
    0 & 0 & 0
  \end{bmatrix}
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
,
 \begin{bmatrix}
   0 & 0 & 0 \\
   0 & 0 & 0 \\
   1 & 0 & 0
 \end{bmatrix}
\odot
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
,
 \begin{bmatrix}
  0 & 0 & 0 \\
  0 & 0 & 0 \\
  0 & 1 & 0
 \end{bmatrix}
\odot
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix}
\Big\rangle
\]
which is equal to
\[
\Big\langle
\begin{bmatrix}
0 \\
1 \\
0
\end{bmatrix},~
\begin{bmatrix}
1 \\
0 \\
0
\end{bmatrix},~
\begin{bmatrix}
0 \\
0 \\
0
\end{bmatrix}
\Big\rangle
\]
This means node $(2,t_1)$ and $(1,t_2)$ are the forward-neighbors of $(1,t_1)$.

Let $(A^{[i]})^Tb = c_i$.
At the second step, we need to compute the following components
\begin{align}
  \label{eq:5}
 \big\langle (A^{[1]})^Tc_1,   (A^{[2]})^T\odot c_1, \ldots,  (A^{[n]})^T\odot c_1  \big\rangle  \\
\big\langle (A^{[2]})^Tc_{2}, \ldots, (A^{[n]})^T\odot c_{2} \big\rangle \\
 \ldots  \\
  \big\langle (A^{[n]})^Tc_n \big\rangle
\label{eq:6}
\end{align}
If we sum these resulting vectors according to time stamps, we obtain
the forward-neighbors of the nodes computed at step $1$.
For $n =2$, we have
\[
\Big\langle
(A^{[1]})^Tc_1,~
(A^{[2]})^T\odot c_1 + (A^{[2]})^T c_2
\Big\rangle.
\]
If we form a block triangular matrix $\bm{A_n}^T$
\[
\bm{A_n}^T =
\begin{bmatrix}
(A^{[1]})^T &   0            & \cdots  &  & 0 \\
\sigma(A^{[2]})^T & (A^{[2]})^T & 0         &\cdots  & \vdots \\
\sigma(A^{[3]})^T & \sigma(A^{[3]})^T & (A^{[3]})^T & \cdots \\
   \vdots              &                              &      \vdots            &  &  \vdots \\
\sigma(A^{[n]})^T &                  &       \cdots                      &  & (A^{[n]})^T
\end{bmatrix},
\]
where  $\sigma(A^{[i]})^Tb = (A^{[i]})^T\odot b$, and a block vector $\bm{b}$
\[
\bm{b}^T = [b^T, 0, \cdots, 0  ],
\]
then \eqref{eq:out_ne} is equivalent to $\bm{A_n}^T\bm{b}$ and
summing \eqref{eq:5}-\eqref{eq:6} is equivalent to $\bm{A_n}^T(\bm{A_n}^T\bm{b})$.
In general, we can define an evolving-graph BFS using
block vector-matrix multiplications. The algorithm is presented in Algorithm
\ref{alg:bfsla}.

@@

% adjacency matrices
Suppose $G_n$ is represented by adjacency lists.
Then the number of adjacency lists is the same as the number
of active nodes in $G_n$. So for large $n$, this representation
can use a large amount of memory.
For the graph $G^{[i]} = (V^{[i]}, E^{[i]})$ at time stamp $i$,
we denote the active node set of $V^{[i]}$ as $\tilde V^{[i]}$,
where we have $\tilde V^{[i]} \subseteq V^{[i]}$.
Step $10$ of  Algorithm \ref{alg:bfs} ensures each active node-time stamp pair in $G_n$ is enqueued in $next$ (and then $frontier$) at most once.
And so \texttt{forward-neighbors} loops through each
active node-time stamp pair at most once.
The operations of enqueuing take $O(1)$ time.
Thus the total operations of enqueuing are
$O(\sum_{i=1}^n|\tilde V^{[i]}|)$, where $|\cdot|$ represents the number of elements of a set.  Let $\sum_i |\tilde V^{[i]}| = |\tilde V|$ and $\sum_i |E^{[i]}| = |\hat E|$.
Then the time of scanning
\texttt{forward-neighbors} is
\[
\sum_{i=1}^n\sum_{v \in \tilde V^{[i]}} \texttt{forward-neighbors}(v) =
O(|\hat E|),
\]
by the definition of forward-neighbors.
Hence the total running time is $O(|\hat E|+ |\tilde V|)$.

@@

Given a root node $(s, t)$, during the execution the BFS discussed above builds a search tree, denoted by $T(s,t)$,
where nodes are temporally connected.
\begin{lemma}
 For a node active at different time stamps, say $(s, t_1)$ and $(s,t_2)$ such that $t_1 < t_2$,  we have $T(s, t_2) \subset T(s, t_1)$.
\end{lemma}
\proof
Suppose $(v, t)$ belongs to $T(s, t_2)$. Then there exits an temporal path
(constructed by the evolving graph BFS) from $(s,t_2)$ to $(v,t)$. Since both
$(s,t_1)$ and $(s,t_2)$ are active, $(s,t_2)$ is an out-neighbor of $(s,t_1)$ by definition, i.e., there exits an edge from $(s,t_1)$ to $(s,t_2)$.  Hence there exits an temporal path
from $(s,t_1)$ to $(v,t)$ and so $(v,t) \in T(s,t_1)$.
\endproof
